The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor Couette flow

### 2016-10-03 paper/thingap_mri_paper.tex
IVP/GLE text

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/thingap_streamfuncs_velocity_Pm_1E-3.png}%{allorders_velocity.eps}
\caption{First order (left), second order (center), and total (right) velocity perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to faster speeds. Colorbar represents azimuthal velocity.}\label{allorders_velocity}
\end{figure*}

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/thingap_streamfuncs_Bfield_Pm_1E-3.png}
\caption{First order (left), second order (center), and total (right) magnetic field perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to higher magnetic field strengths. Colorbar represents azimuthal magnetic field strength.}\label{allorders_Bfield}
\end{figure*}

@@

Figure~\ref{fig:b_saturation} shows $B_z$ and the terms corresponding to steady state inductive balance (equation~(\ref{eqset2} with $\partial_t B_{y} = 0$). Here, the instability acts to push magnetic field toward the boundaries in both the bulk and the boundary layers. Note also that the total magnetic energy in the $B_z$ component is significantly larger than the initial condition. \citet{Ebrahimi:2009} considered the saturation of a single, strongly super-critical MRI mode allowed to interact non-linearly only with itself and the mean. They considered two important cases, one in which the mean flow was forced to remain at its initial, quasi-Keplerian state for all time, and one in which the background flow was allowed to evolve. This is a crucial difference between the shearing box and our narrow-gap TC flow: perturbations in our simulation can adjust the background flow, whereas in a shearing box, the shear periodicity forbids perturbations from affecting the mean flow. In the case with a freely evolving background flow, \citet{Ebrahimi:2009ey} found a saturated state quite similar to ours: field pushed to the boundaries, and a reduction in shear in the bulk of the flow. Their flows have less pronounced boundary layers, likely because of their much larger $\Pm = 0.1 -1$.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/IVP_plot_thingap_AA_phase_2lambdacrit_2.png}%{allorders_velocity.eps}
\caption{IVP for two critical wavelengths.}\label{fig:IVP_2lambdacrit}
\end{figure*}

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/IVP_plot_thingap_AA_phase_2lambdacrit_10.png}%{allorders_velocity.eps}
\caption{IVP for 10 critical wavelengths.}\label{fig:IVP_10lambdacrit}
\end{figure*}

>>>>

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../figures/thingap_streamfuncs_velocity_Pm_1E-3.png}%{allorders_velocity.eps}
\caption{First order (left), second order (center), and total (right) velocity perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to faster speeds. Colorbar represents azimuthal velocity.}\label{allorders_velocity}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../figures/thingap_streamfuncs_Bfield_Pm_1E-3.png}
\caption{First order (left), second order (center), and total (right) magnetic field perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to higher magnetic field strengths. Colorbar represents azimuthal magnetic field strength.}\label{allorders_Bfield}
\end{figure*}

@@

To close the perturbation series we enforce a solvability criterion on Equation \ref{eq:ordere3} (see Appendix \ref{app:matrices}). This leads to an amplitude equation for $\alpha(T, Z)$ that governs the slow length- and timescale evolution of the system. This amplitude equation is

\beq
\label{eq:gle}
\partial_T \alpha = b \alpha + d \partial_Z^2 \alpha - c \alpha \left|\alpha^2\right|,
\eeq

a real Ginzburg-Landau equation.

@@

Figure~\ref{fig:b_saturation} shows $B_z$ and the terms corresponding to steady state inductive balance (Equation~\ref{eqset2} with $\partial_t B_{y} = 0$). Here, the instability acts to push magnetic field toward the boundaries in both the bulk and the boundary layers. Note also that the total magnetic energy in the $B_z$ component is significantly larger than the initial condition. \citet{Ebrahimi:2009} considered the saturation of a single, strongly super-critical MRI mode allowed to interact non-linearly only with itself and the mean. They considered two important cases, one in which the mean flow was forced to remain at its initial, quasi-Keplerian state for all time, and one in which the background flow was allowed to evolve. This is a crucial difference between the shearing box and our narrow-gap TC flow: perturbations in our simulation can adjust the background flow, whereas in a shearing box, the shear periodicity forbids perturbations from affecting the mean flow. In the case with a freely evolving background flow, \citet{Ebrahimi:2009ey} found a saturated state quite similar to ours: field pushed to the boundaries, and a reduction in shear in the bulk of the flow. Their flows have less pronounced boundary layers, likely because of their much larger $\Pm = 0.1 -1$.

@@

The real Ginzburg-Landau equation describes the amplitude behavior of our system close to threshold. Although the form of the equation is generic, its coefficients depend on the specific physics of our system and govern its detailed evolution (see Appendix \ref{app:matrices}). We simulate the evolution of the MRI amplitude equation by solving Equation \ref{eq:gle} on a Fourier basis in $Z$ using Dedalus. We initialize uniform random noise of amplitude $-10^{-3}$ to $+10^{-3}$ in $Z$, and timestep the system using a four-stage, fourth-order Runge-Kutta integrator. We evolve the system for $100 \Omega_0^{-1}$ in timesteps of $0.02 \Omega_0^{-1}$. Results are shown in Figures \ref{fig:IVP_2lambdacrit} and \ref{fig:IVP_10lambdacrit}. The system quickly organizes itself into rolls in $Z$ bounded by the analytic saturation amplitude $\alpha_s = \sqrt{b/c}$. The specific geometry depends on the number of critical wavelengths $\lambda_{crit} = 2\pi/k_c$ that are initialized in $Z$. Figure \ref{fig:IVP_2lambdacrit} shows that a system with a height equal to two critical wavelengths will be modulated by simple rolls of sinusoidal amplitude. The saturation amplitude pattern becomes more complicated when more modes are allowed to interact. Figure \ref{fig:IVP_10lambdacrit} shows the evolution of a system of height $10 \lambda_{crit}$. While still bounded by $\alpha_s$, the saturation amplitude exhibits a nonlinear phase geometry due to the nonlinear interaction of modes in $Z$.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/IVP_plot_thingap_AA_phase_2lambdacrit_2.png}%{allorders_velocity.eps}
\caption{Evolution of the Ginzburg-Landau amplitude equation (Equation \ref{eq:gle}) on a Fourier $Z$ domain of length $2 \lambda_{crit}$, where $\lambda_{crit} = 2\pi/k_c$ is the critical wavelength of the system. Top left panel shows the evolution of the amplitude observable $\alpha \alpha^*$ on the full $Z$ domain as a function of time $T$. Bottom left panel shows the amplitude $alpha$ at the final timestep shown, where the black line is the real part $\mathrm{Re}\{\alpha(t = t_{final})\}$ and the gray line is the imaginary part $\mathrm{Im}\{\alpha(t = t_{final})\}$. The final amplitude is bounded by the analytic saturation amplitude $\alpha_{saturation} = \pm \sqrt{b/c}$. Top right panel shows the evolution of the phase angle $\phi = \mathrm{arctan} (\mathrm{Im}(\alpha)/\mathrm{Re}(\alpha))$ on the same domain. Bottom panel shows the phase angle as a function of $Z$ for the final timestep. Note that the phase angle is wrapped on a $2\pi$ domain, such that $\pi$ = $-\pi$, as indicated by the circular colorbar.}\label{fig:IVP_2lambdacrit}
%\end{figure*}

%\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/IVP_plot_thingap_AA_phase_2lambdacrit_10.png}%{allorders_velocity.eps}
\caption{Same as Figure \ref{fig:IVP_2lambdacrit} but for a $Z$ domain of length $10 \lambda_{crit}$.}\label{fig:IVP_10lambdacrit}
\end{figure*}
